Investigating the alternative prey hypothesis with the POMP framework

### 2024-04-22 draft.Rmd

- **year**: Self explanatory
- **CPUE**: Catch per unit effort, expressed as a number of birds shot per hunter per year.
- **log.CPUE**: Logarithm transformation of CPUE
- **peak_rodent_year**: Peak rodent year is scored as “yes”, otherwise “no”. Occurrence of peak
rodent years were extracted from four sources: 1871-1949 (Wildhagen 1952), 1932-1971
(Myrberget 1982a), 1971-1979 (Christiansen 1983), 1981-1985 (Frafjord 1988), 1986- 1989
(Selås et al. 2013), and 1990-2013 (Framstad 2017). Note that 1980 is missing as we were
not able to find an estimate for this year.
- **NAO**: The mean North Atlantic Oscillation index for the months May through
July.

```{r libraries, echo=FALSE, message=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(forecast)
library(dplyr)
library(cowplot)
library(pomp)
```

```{r data_table, echo=FALSE}
data <- read_excel("./data/bird_data.xlsx")
data |>
  rename(peak_rodent_year = "peak rodent year") |>
  rename(NAO = "NAO May_Jun_Jul")-> data

knitr::kable(data[1:5, 1:7], caption = "Data")
```

@@

p3 <- ggplot(data, aes(x=year, y=log.CPUE)) + 
  geom_line(color="steelblue") + 
  geom_point() + 
  geom_rect(
    data = subset(data, peak_rodent_year == 'yes'),
            aes(xmin = year - 0.5, xmax = year + 0.5, ymin = -Inf, ymax = Inf),
            fill = "blue", alpha = 0.3
    ) + 
  theme_light() + 
  xlab("Year") + 
  stat_smooth(
    color = "#FC4E07",
    method = "loess"
  )

@@

We will start off by fitting a simple ARMA model with parameters chosen using the algorithmic approach, grid search. This simple model will serve as a reference against which the performance of the more advanced POMP model is measured. Essentially, the ARMA model will serve as a yardstick for managing expectations. Of the models shown in (table #), we select the one with the Akaike’s information criterion (AIC). AIC is essentially “minus twice the maximized log likelihood plus twice the number of parameters [cite],” and is defined by:

@@

```{r, echo=FALSE}
# This code snippet is to add a new column that mutates peak_rodent_year into 
# binary values for the POMP model.
data |>
  mutate(rodent_binary = ifelse(peak_rodent_year == 'yes', 1, 0)) -> data
```

@@

The measurement model $Y(t)$ is our ptarmigan count proxy, *log.CPUE*, $y(t) =Negative\ Binomial(mean=\rho\beta_t, \sigma)$, and the process model $X(t)$ includes predator and alternative prey (rodent) count.

@@

- $R(t)=$ peak rodent year, from data

@@

```{r, eval=FALSE}
Csnippet("
  double dNf  = rbinom(F, 1-exp);
  double dF_p = rbinom(F, 1-exp(rho_t * beta_t * dt));
  double dF_m = rbinom(F, 1-exp(theta*dt));

  B = 
  F += dF_p - dF_m;

") -> step

Csnippet("
  B = 
  F = 
") -> rinit

Csnippet("

") -> rproc

Csnippet("
  count = dnbinom(rho*B, sigma)
") -> rmeas

Cnsippet("

") -> dmeas

mod <- pomp(rprocess = euler(step, delta.t =),
            times = "year", t0 = 1872,
            rinit = rinit,
            rprocess = eurler(step, delta.t=),
            rmeasure = rmeas, 
            dmeasure = , 
            paramnames = c("rho", "theta", "delta"),

)
```

>>>>

- **year**: Self explanatory
- **log.CPUE**: Logarithm transformation of CPUE, catch per unit effort, expressed as a number of birds shot per hunter per year.
- **peak_rodent_year**: Peak rodent year is scored as “yes”, otherwise “no”. Occurrence of peak
rodent years were extracted from four sources: 1871-1949 (Wildhagen 1952), 1932-1971
(Myrberget 1982a), 1971-1979 (Christiansen 1983), 1981-1985 (Frafjord 1988), 1986- 1989
(Selås et al. 2013), and 1990-2013 (Framstad 2017).

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(forecast)
library(dplyr)
library(cowplot)
library(pomp)
```

```{r data_table, echo=FALSE}
data <- read_excel("./data/bird_data.xlsx")
data |>
  rename(R = "peak rodent year") |>
  rename(NAO = "NAO May_Jun_Jul") |>
  rename(logCPUE = "log.CPUE") |>
  mutate(R = ifelse(R == "no", 0, 1)) -> data

subset <- c("year", "logCPUE", "R")

knitr::kable(data[subset], caption = "Data")
```

@@

p3 <- ggplot(data, aes(x=year, y=log.CPUE)) + 
  geom_line(color="steelblue") + 
  geom_point() + 
  geom_rect(
    data = subset(data, peak_rodent_year == 1),
            aes(xmin = year - 0.5, xmax = year + 0.5, ymin = -Inf, ymax = Inf),
            fill = "blue", alpha = 0.3
    ) + 
  theme_light() + 
  xlab("Year") + 
  stat_smooth(
    color = "#FC4E07",
    method = "loess"
  )

@@

We will start off by fitting a simple ARMA model with parameters chosen using the algorithmic approach, grid search. This simple model will serve as a reference against which the performance of the more advanced POMP model is measured. In theory, the ARMA model will serve as a yardstick for managing expectations. Of the models shown in (table #), we select the one with the Akaike’s information criterion (AIC). AIC is essentially “minus twice the maximized log likelihood plus twice the number of parameters [cite],” and is defined by:

@@

@@

The measurement model $Y(t)$ is our ptarmigan count proxy, *log.CPUE*, $y(t) =Negative\ Binomial(mean=\rho\beta_t, \sigma)$, and the process model $X(t)$ includes predator (fox), preferred prey(rodent) and alternative prey (bird) count.

@@

- $Z(t)= R(t)$ peak rodent yea

$$Y_n = N$$

@@

```{r}
statenames <- c("logF", "logB")
paramnames <- c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB", 
                "logF_0", "logB_0", "logRho", "sigma_obs")
obs_names <- c("logCPUE")
covarnames <- c("R")
# logF_0 = 0 normalizing to assume unit initial population
# sigmaF roughly the noise in a year
param_guess <- c(
  logF_0 = 0, 
  logB_0 = 0,
  a = 1,
  b = 2, 
  c = 1,
  alpha = 1, 
  Beta = 1,
  gamma = 0.5,
  sigmaF = 0.1, 
  sigmaB = 0.1, 
  logRho = 3, 
  sigma_obs = 0.1
)
```

```{r}
bird <- data[subset]
head(bird)
```

```{r, eval=FALSE}
Csnippet("
  double dwW, dwB;

  dwF = rgammawn(sigmaF, dt);
  dwB = rgammawn(sigmaB, dt);

  logF += (b*R + c*exp(logB)*(1-gamma*R)-a)*dwF;
  logB += (alpha - Beta*exp(logF)*(1-gamma*R))*dwB;
") -> step

Csnippet("
  logF = logF_0;
  logB = logB_0;
") -> rinit

Csnippet("
  log.CPUE = rnorm(logB + logrho, signma_obs)
") -> rmeas

Csnippet("
  lik = dnorm(logCPUE, lobB + logrho, sigma_obs.give_log)
") -> dmeas

bird |>
  pomp(times="year", t0=1872,
       rprocess=euler(step, delta.t=1/52),
       rinit=rinit,
       rmeasure=rmeas,
       dmeasure=dmeas,
       statenames=statenames,
       paramnames=paramnames,
       obs_names <- c("logCPUE"),
       covarnames <- c("R")
    ) -> mod

```

- `rinit`: simulator for the initial-state distribution, i.e., the distribution of the latent state at time $t_0$
